A Type System for Safe and Scalable Parallelism using Logical Regions \\ (Extended Version)

### 2012-07-19 doc/popl2013full/langdef.tex
big proof

\noindent Two additional properties are proven for each subexpression:
\begin{itemize}
\item $\valueconsist{v}{T}$, the expression yields a value of the right type
\item $\traceconsist{E}$, values written to the store match the heap types of their locations
\end{itemize}
Figure~\ref{fig:constprop} gives the precise definition of these properties.

@@

The only thing remaining to show is $\valueconsist{v'}{T_1}$.
From [E-Pack] we have   $v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle$.
From [T-Pack] we have $T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1$
and $\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^*$.  
%
% FIXME  The next sentence can't be made fully formal with the current setup.
%
The result
follows from the definition of value consistency on region relationships.

@@

Therefore, we can conclude by induction
$\valueconsist[M']{v}{T}$ and
$\traceconsist{E_2}$ and 
$\privconsist{E_2}{\Phi}$.
Because $T$ may not mention any of the regions introduced in the partitioning, we have
$\valueconsist{v}{T}$.
From [E-Part] we have $$valid\_interleave(S, C, E', E_1, E_2)$$ and using Lemma~\ref{lemma:heapconst:effects2} we
prove $\traceconsist{E'}$.
To finish the case, we note that from $\privconsist{E_1}{\Phi}$, $\privconsist{E_2}{\Phi}$, 
$valid\_interleave(S, C, E', E_1, E_2)$, and using Lemma~\ref{lemma:effsound:effects2} we can conclude $\privconsist{E'}{\Phi}$.

@@

For subexpression $e_1$, from [E-Call] we have $\opsenv{e_1}{v_1}{E_1}$ and from [T-Call] we have
$\typeenv{e_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$.
By induction, 
$\valueconsist{v_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$,
$\traceconsist{E_1}$ and 
$\privconsist{E_1}{\Phi}$.

For the other subexpressions that are arguments to the call the reasoning is very similar; we show
only the last argument $e_n$.  First, 
$$S_n = \text{apply}(E_{n-1},\text{apply}(E_{n-2},\ldots \text{apply}(E_1,S) \ldots )$$
From Lemma~\ref{lemma:heapconst:apply} it follows that $\storeconsist{S_n}{H}$.
We also have $\opsenv{e_n}{v_n}{E_n}$ from [E-Call] and
$\typeenv{e_n}{T_n[r_1/r'_1,\ldots,r_k/r'_k]}$ from [T-Call].
By induction, 
$\valueconsist{v_n}{T_n[r_1/r'_1,\ldots,r_k/r'_k]}$,
$\traceconsist{E_n}$ and 
$\privconsist{E_n}{\Phi}$.

Now, from [E-Call] we have
$valid\_interleave(S, C, E', \oton{E})$
and furthermore
$M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \}$  and
$L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \}$ and
$S' = \text{apply}(S, E')$ and
$\opsenv[M=M',L=L',S=S',C=C']{e_{n+1}}{v_{n+1}}{E_{n+1}}$.
From [T-Program] we know
$\typeenv[G=\Gamma',O=\emptyset,P=\Phi']{e_{n+1}}{T_r}$,
where $e_{n+1}$ is the body of the function.
To apply the induction hypothesis to $e_{n+1}$ we must
show 
\begin{itemize}
\item $\mapconsist[M']{\emptyset}$,
which is immediate.

\item $\localconsist[M']{L'}{\Gamma'}$ where
$\Gamma'(a_i) = T_i$and $\storeconsist{S'}{H}$.
\end{itemize}

$\Gamma(id) = \forall r'_1, \ldots r'_k.(\oton{T}),\Phi', Q' \rightarrow T_r$ 
$\text{function }id[\otok{k}{r'}](a_1 : T_1, \ldots, a_n : T_n), \Phi', Q' : T_r = e_{n+1}$

$E'_{n+1} = mark\_coherence(E_{n+1}, M' \llbracket Q' \rrbracket)$ 
$valid\_interleave(S, C, E'', E', E_{n+1})$

$\typeenv{e_i}{T_i[r_1/r'_1,\ldots,r_k/r'_k]}$
$\Phi'[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*$

\end{itemize}
\end{proof}

@@

\begin{proof}
The proof is by contradiction.  Assume $E_1 \cancel\nonint E_2$.  Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \cancel\nonint \epsilon_2$.  This requires that
$\epsilon_1$ and $\epsilon_2$ be operations to the same location $l$ and not both be reads or both
reductions using the same function name.  Assume $\epsilon_1 = read(l, c_1, v_1)$ and $\epsilon_2 = write(l, c_2, v_2)$.  The definition of $\privconsist{E_1}{\Phi_1}$ tells us that there must be some $r_1$ satisfying $l \in M(r_1) \wedge reads(r_1) \in \Phi_1$.  Similarly, there must be some $r_2$ satisfying $l \in M(r_2) \wedge writes(r_2) \in \Phi_2$.  Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 \cancel{\nonint[M]} \phi_2$ and $\Phi_1 \cancel{\nonint[M]} \Phi_2$.  There are seven other cases to consider (using $id_1 \not= id_2$ for the reduce-reduce case), but all yield the same result.
\end{proof}

>>>>

\noindent Two additional properties are proven for each subexpression:
\begin{itemize}
\item $\valueconsist{v}{T}$, the expression yields a value of the right type
\item $\traceconsist{E}$, values written to the store match the heap types of their locations
\end{itemize}
Figure~\ref{fig:constprop} gives the precise definition of these
properties.\footnote{Note that in the definition of
$\mapconsist{\Omega}$, the quantified variables $r_i$ and $r_j$
range over both logical and concrete regions.  Since $M$'s domain is
logical regions, we extend $M$ to constraints with mixed logical and
concrete regions in the standard way by defining
$M \llbracket \rho \rrbracket = \rho$.}

@@

The only thing remaining to show is $\valueconsist{v'}{T_1}$.
From [E-Pack] we have   $v' = \langle \langle \otok{k}{\rho}, v \rangle \rangle$.
From [T-Pack] we have $T_1 = \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1$
and $\Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Omega^*$.

Observe that 
$$M \llbracket \exists r'_1, \ldots r'_n.\ T_2\text{ where }\Omega_1 \rrbracket$$
replaces all free region names in $T_2$ and particularly $\Omega_1$ by concrete regions, so
that the only region names remaining are $r'_1,\ldots,r'_n$.  It is then easy to show that
$$\mapconsist[\{(r'_i,\rho_i)\}]{M \backslash \{r'_1,\ldots, r'_n \}\llbracket \Omega_1 \rrbracket}$$
from which $\valueconsist{v'}{T_1}$ follows (see Figure~\ref{fig:constprop}).

@@

Therefore, we can conclude by induction
$\valueconsist[M']{v}{T}$ and
$\traceconsist{E_2}$ and 
$\privconsist{E_2}{\Phi}$.
Because $T$ may not mention any of the regions introduced in the partitioning, we have
$\valueconsist{v}{T}$.
From [E-Part] we have $$valid\_interleave(S, C, E', E_1, E_2)$$ and using Lemma~\ref{lemma:heapconst:effects2} we
prove $\traceconsist{E'}$.
To finish the case, we note that from $\privconsist{E_1}{\Phi}$, $\privconsist{E_2}{\Phi}$, 
$$valid\_interleave(S, C, E', E_1, E_2)$$, and using Lemma~\ref{lemma:effsound:effects2} we can conclude $\privconsist{E'}{\Phi}$.

@@

For subexpression $e_1$, from [E-Call] we have $$\opsenv{e_1}{v_1}{E_1}$$ and from [T-Call] we have
$$\typeenv{e_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$$
By induction, 
$\valueconsist{v_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$,
$\traceconsist{E_1}$ and 
$\privconsist{E_1}{\Phi}$.

For the other subexpressions that are arguments to the call the reasoning is very similar; we show
only the last argument $e_n$.  First, 
$$S_n
= \text{apply}(E_{n-1},\text{apply}(E_{n-2},\ldots \text{apply}(E_1,S) \ldots
)$$ An induction on the number of applications of {\em apply} using
Lemma~\ref{lemma:heapconst:apply} shows that
$\storeconsist{S_n}{H}$.  We also have $$\opsenv{e_n}{v_n}{E_n}$$ from
[E-Call] and $$\typeenv{e_n}{T_n[r_1/r'_1,\ldots,r_k/r'_k]}$$ from
[T-Call].  By induction,
$\valueconsist{v_n}{T_n[r_1/r'_1,\ldots,r_k/r'_k]}$,
$\traceconsist{E_n}$ and $\privconsist{E_n}{\Phi}$.

Now, from [E-Call] we have
$$valid\_interleave(S, C, E', \oton{E})$$
and furthermore
$$M' = \{ (r'_1, M(r_1)), \ldots (r'_k, M(r_k)) \}$$
$$L' = \{ (a_1, v_1), \ldots, (a_n, v_n) \}$$
$$S' = \text{apply}(S, E')$$
We also know from [E-Call] that
$$\opsenv[M=M',L=L',S=S',C=C']{e_{n+1}}{v_{n+1}}{E_{n+1}}$$
where $e_{n+1}$ is the body of the called function.
From [T-Program] we know
$$\typeenv[G=\Gamma',O=\emptyset,P=\Phi']{e_{n+1}}{T_r}$$
where $\Gamma'(a_i) = T_i$.
To apply the induction hypothesis to $e_{n+1}$ we want to show
\begin{enumerate}
\item $\storeconsist{S'}{H}$, 
\item $\mapconsist[M']{\emptyset}$, and
\item $\localconsist[M']{L'}{\Gamma'}$ 
\end{enumerate}

For (1), using $\traceconsist{E_i}$ for $1 \leq i \leq n$ and applying Lemma~\ref{lemma:heapconst:effects2} $n - 1$ times
we can conclude that $\traceconsist{E'}$.  Then because $\storeconsist{S}{H}$ and $S' = \text{apply}(S, E')$ by
Lemma~\ref{lemma:heapconst:apply} we have $\storeconsist{S'}{H}$.

Part (2) is immediate by the definition of mapping consistency.

For (3), in previous steps we have shown  $$\valueconsist{v_i}{T_i[r_1/r'_1,\ldots,r_k/r'_k]}$$ for $1 \leq i \leq n$.
Since $M \llbracket r_i \rrbracket = M' \llbracket r'_i \rrbracket$ it follows that
$M \llbracket T_i[r_1/r'_1,\ldots,r_k/r'_k \rrbracket = M' \llbracket T_i \rrbracket$ and therefore
$\valueconsist[M']{v_i}{T_i}$, from which the result follows.

Thus, by induction, we conclude that
$\valueconsist[M']{v_{n+1}}{T_r}$, $\traceconsist{E_{n+1}}$ (under
$M'$) and $\privconsist[M']{E_{n+1}}{\Phi'}$.  Because 
$$M \llbracket T_r[r_1/r'_1,\ldots,r_k/r'_k \rrbracket = M' \llbracket T_r \rrbracket$$ we can conclude
$\valueconsist{v_{n+1}}{T_r[r_1/r'_1,\ldots,r_k/r'_k}$.

To show that
$\traceconsist{E_{n+1}}$ under $M'$ implies $\traceconsist{E_{n+1}}$
under $M$, note that $H(\cdot)$ does not depend on the mapping (because it includes only concrete, not logical, regions) so the
types of locations are unaffected by the mapping (see
Figure~\ref{fig:constprop}).  Now, the environment
$M' \llbracket \Gamma' \rrbracket$ is used to look up reduction
functions, but the reduction function signatures are restricted to
have no regions named in their arguments and so
$$M \llbracket \Gamma' \rrbracket(id)  = M' \llbracket \Gamma' \rrbracket(id)$$
Finally, from 
$$valid\_interleave(S, C, E'', E', E_{n+1})$$
and Lemma~\ref{lemma:heapconst:effects2} we conclude $\traceconsist{E''}$.

For the last consistency property of the function call, we first observe
that $\privconsist[M']{E_{n+1}}{\Phi'}$ is equivalent to 
$$\privconsist{E_{n+1}}{\Phi[r_1/r'_1,\ldots,r_k/r'_k]}$$
because, again, $E_{n+1}$ does not mention logical region names and
$M' \llbracket r' \rrbracket = M \llbracket r \rrbracket$.
From [T-Call] we have 
$$\Phi[r_1/r'_1,\ldots,r_k/r'_k] \subseteq \Phi^*$$
and so
$\privconsist{E_{n+1}}{\Phi^*}$.  Applying Lemma~\ref{lemma:phiclosure} we conclude
that $\privconsist{E_{n+1}}{\Phi}$.
Finally, from Lemma~\ref{lemma:effsound:effects2} we conclude $\privconsist{E''}{\Phi}$.
\end{itemize}
\end{proof}

@@

\begin{proof}
The proof is by contradiction.  Assume $E_1 \nonint E_2$.  Then there must be some $\epsilon_1$ in $E_1$ and some $\epsilon_2$ in $E_2$ such that $\epsilon_1 \nonint \epsilon_2$.  This requires that
$\epsilon_1$ and $\epsilon_2$ be operations to the same location $l$ and not both be reads or both
reductions using the same function name.  Assume $\epsilon_1 = read(l, c_1, v_1)$ and $\epsilon_2 = write(l, c_2, v_2)$.  The definition of $\privconsist{E_1}{\Phi_1}$ tells us that there must be some $r_1$ satisfying $l \in M(r_1) \wedge reads(r_1) \in \Phi_1$.  Similarly, there must be some $r_2$ satisfying $l \in M(r_2) \wedge writes(r_2) \in \Phi_2$.  Letting $\phi_1 = reads(r_1)$ and $\phi_2 = writes(r_2)$, we have $M(r_1) \cap M(r_2) \not= \emptyset$ (it contains at least $l$), so $\phi_1 {\nonint[M]} \phi_2$ and $\Phi_1 {\nonint[M]} \Phi_2$.  There are seven other cases to consider (using $id_1 \not= id_2$ for the reduce-reduce case), but all yield the same result.
\end{proof}
